

```{r}

benguela <- read.csv("http://faraway.neu.edu/data/assn3_benguela.csv")
california <- read.csv("http://faraway.neu.edu/data/assn3_california.csv")
canary <- read.csv("http://faraway.neu.edu/data/assn3_canary.csv")
humboldt <- read.csv("http://faraway.neu.edu/data/assn3_humboldt.csv")

#create column with before and after
beng<-ifelse(benguela$year<=2024, "before", "after")
cali<-ifelse(california$year<=2024, "before", "after")
cana<-ifelse(canary$year<=2024, "before", "after")
humb<-ifelse(humboldt$year<=2024, "before", "after")

#adding new column to main tables (redefining)
benguela<- cbind(benguela, beng)
california<-cbind(california, cali)
canary<-cbind(canary, cana)
humboldt<-cbind(humboldt, humb)
```


### Question 2

```{r}
beng.means<-rowMeans(benguela[ ,1:22], na.rm=FALSE)
benguela2<-data.frame(subset(benguela, select="year"), beng, beng.means)

cali.means<-rowMeans(california[ ,1:22], na.rm=FALSE)
california2<-data.frame(subset(california, select="year"), cali, cali.means)

cana.means<-rowMeans(canary[ ,1:22], na.rm=FALSE)
canary2<-data.frame(subset(canary, select="year"), cana, cana.means)

humb.means<-rowMeans(humboldt[ ,1:22], na.rm=FALSE)
humboldt2<-data.frame(subset(humboldt, select="year"), humb, humb.means)
```

### Question 3

```{r}

## Normal 
beng.before<-subset(benguela2, subset=beng=="before", select=c("year", "beng.means"))
beng.after<-subset(benguela2, subset=beng=="after", select=c("year", "beng.means"))

cali.before<-subset(california2, subset=cali=="before", select=c("year", "cali.means"))
cali.after<-subset(california2, subset=cali=="after", select=c("year", "cali.means"))

cana.before<-subset(canary2, subset=cana=="before", select=c("year", "cana.means"))
cana.after<-subset(canary2, subset=cana=="after", select=c("year", "cana.means"))

cana.before<-subset(canary2, subset=cana=="before", select=c("year", "cana.means"))
cana.after<-subset(canary2, subset=cana=="after", select=c("year", "cana.means"))

humb.before<-subset(humboldt2, subset=humb=="before", select=c("year", "humb.means"))
humb.after<-subset(humboldt2, subset=cana=="after", select=c("year", "humb.means"))


shapiro.test(beng.before$beng.means)  ##Normal
shapiro.test(beng.after$beng.means)   ##Normal
shapiro.test(cali.before$cali.means)  ##Normal
shapiro.test(cali.after$cali.means)   ##Normal
shapiro.test(cana.before$cana.means)  ##Normal
shapiro.test(cana.after$cana.means)   ##Not Normal
shapiro.test(humb.before$humb.means)  ##Normal
shapiro.test(humb.after$humb.means)   ##Not Normal 


```

Non-parametric tests make fewer assumptions about underlying distribution of the data but have low power. The low power means there is a higher chance of type 2 error and rejecting the null hypothesis when it is false. We will use the Wilcoxon test to determine if mean upwelling differs between periods. The test assumes that the data are symmetrically distributed about the median.

### Question 4

```{r}
wilcox.test(beng.before$beng.means, beng.after$beng.means)
wilcox.test(cali.before$cali.means, cali.after$cali.means)
wilcox.test(cana.before$cana.means, cana.after$cana.means)
wilcox.test(humb.before$humb.means, humb.after$humb.means)
```

Ho = There is no difference in mean upwelling between the periods.
Ha = There is a difference in mean upwelling between the periods.

Since all p-values are lower than the alpha=0.05 threshold, we can reject the null hypothesis and conclude that there is a difference in mean upwelling between periods in each EBCS.

### Question 5

```{r}
avg.beng<-aggregate(beng.means ~ beng, data=benguela2, FUN=mean)
avg.beng<-rbind(avg.beng[2, ], avg.beng[1, ])
sderr.beng<-aggregate(beng.means ~ beng, data=benguela2, FUN= function(x) sd(x)/sqrt(length(x)))

avg.cali<-aggregate(cali.means~cali, data=california2, FUN=mean) 
avg.cali<-rbind(avg.cali[2, ], avg.cali[1, ])
sderr.cali<-aggregate(cali.means~cali, data=california2, FUN= function(x) sd(x)/sqrt(length(x)))

avg.cana<-aggregate(cana.means~cana, data=canary2, FUN=mean) 
avg.cana<-rbind(avg.cana[2, ], avg.cana[1, ])
sderr.cana<-aggregate(cana.means~cana, data=canary2, FUN= function(x) sd(x)/sqrt(length(x)))

avg.humb<-aggregate(humb.means~humb, data=humboldt2, FUN=mean)
avg.humb<-rbind(avg.humb[2, ], avg.humb[1, ])
sderr.humb<-aggregate(humb.means~humb, data=humboldt2, FUN= function(x) sd(x)/sqrt(length(x)))


bp1<-barplot(cbind((avg.beng[, 2]),(avg.cali[, 2]),(avg.cana[, 2]),(avg.humb[, 2])), beside = TRUE, names=c("Benguela", "California", "Canary", "Humboldt"), ylim=c(0,1), col=c("red", "blue"), legend.text = c("Before", "After"), main="Multimodel Mean Upwelling")
arrows(x0 = bp1, x1 = bp1, 
       y0 = (cbind((avg.beng[, 2]),(avg.cali[, 2]),(avg.cana[, 2]),(avg.humb[, 2])) - cbind((sderr.beng[, 2]),(sderr.cali[, 2]),(sderr.cana[, 2]),(sderr.humb[, 2]))), 
       y1 = (cbind((avg.beng[, 2]),(avg.cali[, 2]),(avg.cana[, 2]),(avg.humb[, 2])) + cbind((sderr.beng[, 2]),(sderr.cali[, 2]),(sderr.cana[, 2]),(sderr.humb[, 2]))), angle = 90, code = 3, length = 0.1)
box()
```


### Question 6

Based on the Wilcoxon tests above, along with the trends depicted by the boxplot, we can see that there is a difference in upwellings between period; increasing over time for each EBCS.


Problem 2
---------
### Question 1

We have identified two potential tests that could be used to determine whether the group variance differ:

F-test - used to compare the variances of two populations
Assumptions:
- data in each sample are random
- data in each sample is normally-distributed 

Levene's Test - used to determine whether the variance is equal across groups
Assumptions:
- data in each sample are random
- data in each sample is approximately normally-distributed 

### Question 2

```{r}
var.beng<-aggregate(beng.means ~ beng, data=benguela2, FUN=var)
var.beng<-rbind(var.beng[2, ], var.beng[1, ])

var.cali<-aggregate(cali.means~cali, data=california2, FUN=var) 
var.cali<-rbind(var.cali[2, ], var.cali[1, ])

var.cana<-aggregate(cana.means~cana, data=canary2, FUN=var) 
var.cana<-rbind(var.cana[2, ], var.cana[1, ])

var.humb<-aggregate(humb.means~humb, data=humboldt2, FUN=var)
var.humb<-rbind(var.humb[2, ], var.humb[1, ])

#install.packages('car')
library(car, quietly = TRUE)

leveneTest(benguela2$beng.means~benguela2$beng)
leveneTest(california2$cali.means~california2$cali)
leveneTest(canary2$cana.means~canary2$cana)
leveneTest(humboldt2$humb.means~humboldt2$humb)

```

Ho = variance is identical across all groups
Ha = the variance differs across at least 2 of the groups 

The results from the Levene test show that we reject the null and conclude that the multimodel variances are not equal across periods in the California, Canary and Humbodlt EBSC as the p-values are all less than 0.05. We fail to reject the null and conclude that the multimodel variances are equal across time periods in Benguela. 

### Question 3

```{r}
bp1<-barplot(cbind((var.beng[, 2]),(var.cali[, 2]),(var.cana[, 2]),(var.humb[, 2])), beside = TRUE, names=c("Benguela", "California", "Canary", "Humboldt"), ylim=c(0,0.003), col=c("red", "blue"), legend.text = c("Before", "After"), main="Multimodel Upwelling Variance")
box()
```

### Question 4

The statistical analyses of multimodel variance showed that variance did not differ over time in Benguela. The test also showed that the multimodel variances differed over time in the other three EBCS with the variance decreasing over time in California and increasing over time in Humbodlt. 
Based on the statistical analyses of multimodel means using the Wilcoxon tests, along with the trends depicted by the first figure, we observed that there was a difference in multimodel mean upwellings between periods; increasing over time for each EBCS. In comparing the inconsistencies of the patterns in upwelling variance to that of the upwelling mean, we conclude that the Levene test is a more robust statistical method than the non-parametric Wilcoxon test. 

Problem 3
---------
### Question 1

```{r}
# Reshape data from wide to long format:
california.long <- reshape(california, varying = list(1:22),
    direction = "long", timevar = "model", v.names = "upwelling")
canary.long <- reshape(canary, varying = list(1:22), direction = "long",
    timevar = "model", v.names = "upwelling")
humboldt.long <- reshape(humboldt, varying = list(1:22),
    direction = "long", timevar = "model", v.names = "upwelling")
benguela.long <- reshape(benguela, varying = list(1:22),
    direction = "long", timevar = "model", v.names = "upwelling")
```

To measure model agreement we will use the non-parametric alternative to the ANOVA test; the Kruskal Wallis test. The assumptions of this test are that samples are random, and that the data in each group have the same distribution shape. 

### Question 2

```{r}
(beng.kru <- kruskal.test(upwelling ~ model, data = benguela.long))
(cali.kru <- kruskal.test(upwelling ~ model, data = california.long))
(cana.kru <- kruskal.test(upwelling ~ model, data = canary.long)) 
(humb.kru <- kruskal.test(upwelling ~ model, data = humboldt.long))
```

Ho = all group medians are equal to each other.
Ha = at least two groups differ.

According to the results of the Kruskal-Wallis rank sum test, we can reject the null, and conclude that at least two groups medians differ across all EBCS at the alpha=0.05 level. 

### Question 3

```{r}
beng.mod<-aggregate(upwelling~model, data=benguela.long, FUN=mean)
cali.mod<-aggregate(upwelling~model, data=california.long, FUN=mean)
cana.mod<-aggregate(upwelling~model, data=canary.long, FUN=mean)
humb.mod<-aggregate(upwelling~model, data=humboldt.long, FUN=mean)

par(mfrow=c(2,2))
bp2<-barplot(beng.mod$upwelling, names=c(beng.mod$model), main="Benguela", xlab="Model", ylab="Mean Upwelling", ylim=c(0,1.5))
bp3<-barplot(cali.mod$upwelling, names=c(cali.mod$model), main="California",xlab="Model", ylab="Mean Upwelling", ylim=c(0,1.5))
bp4<-barplot(cana.mod$upwelling, names=c(cana.mod$model), main="Canary",xlab="Model", ylab="Mean Upwelling", ylim=c(0,1.5))
bp5<-barplot(humb.mod$upwelling, names=c(humb.mod$model), main="Humboldt",xlab="Model", ylab="Mean Upwelling",ylim=c(0,1.5))
```

According to the figure above plotting model means for each EBCS, the patterns seem to be consistent with the results of the Kruskal Wallice test. The model means differ in all EBCS in the figure, which agrees with the coclusion of the Kruskal Wallice test that at least two groups medians differ across all EBCS at the alpha=0.05 level.

### Question 4

Not all groups in the sample were normally distributed and had pretty extreme outliers that effected distribution shape. This meant that we were forced to use non-parametric tests for some of the statistical analysis as the varying distrbution shapes defied assumptions of more robust test alternatives. 

