

Problem 1
---------
### Question 1

```{r}
#probability of contracting swine flu
pF2 <- 0.003

#probability of not contracting swine flu
p_F2 <- 1 - pF2

# probability of contracting seasonal flu given that s/he has been infected by the swine flu
pF1F2 <- 0.001 

# probability of not contracting seasonal flu given that s/he has been infected by the swine flu
p_F1F2 <- 1 - pF1F2

#probability of contracting seasonal flu given that s/he has not been infected by the swine flu
pF1_F2 <- 0.001

#probability of contracting seasonal flu given that s/he has not been infected by the swine flu
p_F1_F2 <- 1 - pF1_F2

#Finding probability contracting seasonal flu
pF1 <-  (pF1F2 * pF2) + (pF1_F2 * p_F2) 

#Dependecy Test
(dep <- (pF1 * pF2) == (pF1F2 * pF2))

```
If pF1 and pF2 are independent: P(F1 | F2) = pF1F2 x PF2. 
According to the above test - Swine flu and seasonal flu are independent. 

### Question 2
The independence suggests that contracting swine flu does not effect the likelihood of contracting the seasonal flu. This is observed as the probability of contracting the seasonal flu given that the patient has had swine flu is equal to the probability of contracting the seasonal flu given the patient has not had the swine flu. This suggest that the immune response to these two strains are different. If the immune responses were similar then one would expect the probability of contracting the seasonal flu given that the patient has had the swine flu to be less than the probability of contracting the seasonal flu given the patient has not had the swine flu due to antibody recognition. This is however not the case. 

### Question 3

```{r}

## Probability of detecting swine flu given that it was contracted:
pDF2<-0.7

## False detection rates:
pD_F2<-seq(from=0.01, to=0.5, length=100)

## Prevalance of swine flu P(F2)=0.2
pF2.0.2 <-0.2
p_F2.0.2<- 1-pF2.0.2

## Probability contracting swine flu given that it was detected at P(F2)=0.2
pD.0.2 <- (pDF2*pF2.0.2)+(pD_F2*p_F2.0.2)
pF2D.2 <- ((pDF2*pF2.0.2)/pD.0.2)

## Prevalance of swine flu P(F2)=0.4
pF2.0.4 <-0.4
p_F2.0.4<- 1-pF2.0.4

## Probability contracting swine flu given that it was detected at P(F2)=0.4
pD.0.4 <- (pDF2*pF2.0.4)+(pD_F2*p_F2.0.4)
pF2D.4 <- ((pDF2*pF2.0.4)/pD.0.4)

## Prevalance of swine flu P(F2)=0.6
pF2.0.6 <-0.6
p_F2.0.6<- 1-pF2.0.6

## Probability contracting swine flu given that it was detected at P(F2)=0.6
pD.0.6 <- (pDF2*pF2.0.6)+(pD_F2*p_F2.0.6)
pF2D.6 <- ((pDF2*pF2.0.6)/pD.0.6)

## Prevalance of swine flu P(F2)=0.8
pF2.0.8 <-0.8
p_F2.0.8<- 1-pF2.0.8

## Probability contracting swine flu given that it was detected at P(F2)=0.8
pD.0.8 <- (pDF2*pF2.0.8)+(pD_F2*p_F2.0.8)
pF2D.8 <- ((pDF2*pF2.0.8)/pD.0.8)
```

### Question 4

```{r}

plot(pD_F2, pF2D.2, col="red", ylab="P(F2|D)", xlab="False Detection Rate", ylim = c(0,1), main="Probability of Contracting Swine Flu Given That it was Detected vs. False Detection Rate")
points(pD_F2, pF2D.4, col="blue")
points(pD_F2, pF2D.6, col="green")
points(pD_F2, pF2D.8, col="orange")
abline(a=0.5, b=0, lty=2)
legend(x="bottomleft", pch=c(1), col = c("red", "blue","green","orange"), legend = c("P(F2)=0.2", "P(F2)=0.4","P(F2)=0.6","P(F2)=0.8"))
```

The line at P(F2|D) = 0.5 indicates when the test sensitivity is 50% i.e a true positive rate of 50%. 

### Question 5
The figure suggests that as false detection rate increases, the probability of having the Swine flu given that it was detected (true positive) decreases. This effect is consistent across all levels of flu prevalence P(F2). Increased Swine flu prevalence P(F2) mitigates the decreasing effect of false detection rate on the probablity of having the Swine flu given that it was detected. 

### Question 6
The sensitivity of the test is the probability of detecting the Swine flu given that a patient has the Swine flu (True Positive). From a clinical perspective, the figure suggests that the Swine flu test is less effective at low Swine Flu prevalence. When swine flu prevalence P(F2) is 0.2, test sensitivity drops below 50% rendering the test clinically useless.  

### Question 7

```{r}

pF1V1<-0.05
pF2V1<-0.4
pF2V2<-0.05

pV1<-seq(from=0.001, to=1, length=100)
pV2<-1-pV1

## Pevalence at P(F1|V2)=0.2
pF1V2.2<-0.2
pF1.2<-(pF1V1*pV1)+(pF1V2.2*pV2)
pF2.2<-(pF2V1*pV1)+(pF2V2*pV2)

##Total prevalence of flu at P(F1|V2)=0.2
pF1F2.2<-pF1.2+pF2.2

## Pevalence at P(F1|V2)=0.4
pF1V2.4<-0.4
pF1.4<-(pF1V1*pV1)+(pF1V2.4*pV2)
pF2.4<-(pF2V1*pV1)+(pF2V2*pV2)

##Total prevalence of flu at P(F1|V2)=0.4
pF1F2.4<-pF1.4+pF2.4

## Pevalence at P(F1|V2)=0.6
pF1V2.6<-0.6
pF1.6<-(pF1V1*pV1)+(pF1V2.6*pV2)
pF2.6<-(pF2V1*pV1)+(pF2V2*pV2)

##Total prevalence of flu at P(F1|V2)=0.6
pF1F2.6<-pF1.6+pF2.6

## Pevalence at P(F1|V2)=0.8
pF1V2.8<-0.8
pF1.8<-(pF1V1*pV1)+(pF1V2.8*pV2)
pF2.8<-(pF2V1*pV1)+(pF2V2*pV2)

##Total prevalence of flu at P(F1|V2)=0.8
pF1F2.8<-pF1.8+pF2.8

```

### Question 8

```{r}

plot(pV1, pF1F2.2, ylab="Total Prevalence of the Flu", xlab="Prevalence of Vaccination V1", col="red", ylim=c(0,1), main="Total Prevalence of Flu vs. Prevalence of Vaccination 1")
points(pV1, pF1F2.4, col="blue")
points(pV1, pF1F2.6, col="green")
points(pV1, pF1F2.8, col="orange")
legend(x="topright", pch=c(1), col = c("red", "blue","green","orange"), legend = c("P(F1|V2)=0.2", "P(F1|V2)=0.4","P(F1|V2)=0.6","P(F1|V2)=0.8"))

```

### Question 9
The optimal vaccine strategy varies with the effectiveness of Vaccine 2 (V2) on the seasonal flu F1. As seen in the figure, when Vaccine 2 is ineffective in preventing seasonal flu F1 at P(F1|V2)=0.8, then optimizing treatment by increasing prevalence of Vaccine 1 V1 would decrease total prevalence of the flu. So, an optimal treatment strategy would be purely Vaccine 1. Conversely, when Vaccine 2 is effective in preventing seasonal flu F1 at P(F1|V2)=0.2, then increasing prevalence of Vaccine 1 V1 would not decrease total prevalence of the flu, it would increase the total prevalence. Therefore an optimal treatment strategy would be purely Vaccine 2. When Vaccine 2 is effective in preventing seasonal flu F1 at P(F1|V2)=0.4, then increasing prevalence of Vaccine 1 V1 would not change the prevalence of the flu. In conclusion, due to the linear nature of the curves in the above plot, there is no situation where a mixed strategy is beneficial. Purely V1 or V2 strategies are better depending on the effectiveness of V2 on F1.


Problem 2
---------
### Question 1

```{r}
d1 <- read.csv(file = "http://faraway.neu.edu/data/assn1_dataset1.csv")
```

Null Hypothesis (Ho) = There is no association between age and strain of swine flu contracted.
Alternate Hypothesis (Ha) = There is an association between age and the strain of swine flu contracted. 

### Question 2

```{r}
##compute mean age for seasonal and swine flu
mean.age <- aggregate(age~flu, data =d1, FUN=mean)

##compute range of age for seasonal and swine flu 
range.age <- aggregate(age~flu, data =d1, FUN=range)

```

The mean age of individuals is 43.36 and 42.77 years for seasonal flu and swine flu respectively. The age range of individuals affected by the seasonal flu is 1-85 years, and the age range of individuals effected by the swine flu is 17-66 years. This suggests that the distributions center around similar means but the age distribution of individuals with the seasonal flu is broader than that of the age distribution of individuals with the swine flu. 

### Question 3

```{r}
##subset ages for swine and seasonal flu
F1.age<-subset(d1, subset = flu=="seasonal", select = c(age))
F2.age<-subset(d1, subset = flu=="swine", select = c(age))

##plot age distributions
par(mfrow=c(2,1))
hist(F1.age$age, xlab = "Age (Years)", main="Age Distribution for Seasonal Flu", col="red", xlim=c(0,90), ylim=c(0,25))
hist(F2.age$age, xlab = "Age (Years)", main="Age Distribution for Swine Flu", col="blue", xlim=c(0,90), ylim=c(0,15))

```

### Question 4

Based on the figure above seasonal flu and swine flu do not appear to target the same portion of the population. The seasonal flu seems to affect individuals between the ages of 0-20 years and 60-80 years whereas the swine flu seems to affect individuals between the ages of 15-70. 

### Question 5

```{r}

a<-aggregate(age~flu, data=subset(d1,age>=18 & age <=65), FUN=length)

ce<-aggregate(age~flu, data=subset(d1,age<=18 | age >=66), FUN=length)

a.ce<-cbind(a, ce)

##combine results into table 
DF<-data.frame(a.ce[,c(2,4)])
colnames(DF)<-c("Adults", "Children & Elderly")
row.names(DF)<-c("Seasonal Flu", "Swine Flu")

```

### Question 6

```{r}

mosaicplot(DF, col=c("blue","red"), main="Proportion of age groups with Seasonal and Swine Flu")

```

The Null Hypothesis (Ho) was that there is no association between age and strain of swine flu contracted. The Alternate Hypothesis (Ha) was that there is an association between age and the strain of flu contracted. The pattern in the mosiac plot above appears to support our alternate hypothesis in that the seasonal flu predominantley affects children and elderly and the swine flu predominantly affects adults. 

### Question 7

```{r}
chisq.test(DF, correct = FALSE)

```

The p-value of 2.2x10^-16 is smaller than the critcal value alpha of 0.05. With this we can confidently reject the null hypothesis. The p-value is consistent with the alternate hypothesis that there is an association between age and the strain of flu contracted. 

